Hippocampal replay of experience at real-world speeds

### 2021-07-23 Article.tex
Update on Overleaf.

Surprisingly, our results indicate that most events with coherent spatial content are not dominated by the continuous movement dynamic, but instead correspond to trajectories that are stationary or that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first note that most of the SWRs (14,989 of 21,433 or 70\% of classified SWRs) were classified as having only a single dynamic or dynamic mixture (\FIG{5}A, for specific example SWR see \FIG{3}B), whereas SWRs with multiple dynamics or dynamic mixtures (such as those in \FIG{3}A, C, D, E, and F) were less common. Interestingly, the most common category of SWR were those with only stationary-continuous mixtures (8,944 of 21,433 or 42\% of classified SWRs, \FIG{5}A, \FIGSUPP[5]{fig5sup1}). These events contain representations that move at slower speeds (\FIG{5}D)---similar to those speeds experienced by the animal in the environment (\FIG{5}G)---and are sustained, but slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 73 ms vs. continuous 94 ms, \FIG{5}B). Nonetheless, both the slow-moving events and continuous events frequently represent locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 52 cm vs. continuous 43 cm, \FIG{5}C). This indicates that the content of these SWRs, like those that are classified as continuous, do not represent the animal's actual position. We note that roughly a third of these events persisted for the entire duration of the SWR, but the other two thirds included some time where the model was uncertain about the dynamic. This shows the power of our state space decoder because it allows us identify periods of time when the model has high confidence with each SWR, and use these periods to characterize the SWR overall or to focus on specifically for further analysis.

@@

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with slower speeds of less than 1 m/s (<100 cm/s), which is consistent with the speed that rats move through their environment. The next most common category were stationary events \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019} that activated a persistent representation of a single location, while the "classical" replay events---continuous, extended trajectories through space---made up only about 19\% of classified events. Events with slower dynamics tended to persist for approximately the same length of time as events  with faster continuous dynamics (\texttildelow50-100 ms), consistent with a brief activation of a representation of a location or a short trajectory through space. Events with slower dynamics most often represented the animal's current location, similar to continuous events identified in previous work \citep{DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}; however, slower events were also capable of representing locations far from the animal.

@@

Fourth, our model avoided using large temporal bins and does not make assumptions about the linear progression of events. Instead, the model allows for movement in any direction allowed by geometry of the environment, and further, still accounts for the uncertainty of the data. Because the model gives a moment-by-moment estimate of the position and dynamic, this also minimizes the effect of misidentifying the SWR start and end times. Combined with an acausal estimation of latent position, this approach allows us to identify the specific timing of transitions in dynamics at a fine timescale. In addition to our SWR-based analyses, this approach could also be used to decode activity during movement, where we might expect that place-specific activity would express combinations of stationary and continuous dynamics reflecting different speeds of representational movement during each cycle of the theta rhythm. Similarly, because our approach uses the same encoder-decoder approach as the standard decoder, it can be used to investigate sequence dynamics before experience \cite{}.

@@

\subsection*{Classification of Dynamics}
We used a threshold of 0.80 to classify the probability of each state $Pr(I_{k} \mid O_{1:T})$ into 5 categories. Time periods during sharp wave ripples are labeled as stationary, continuous, or fragmented when the probability of each individual state is above 0.80. Time periods are labeled as stationary-continuous-mixture or fragmented-continuous-mixture when the sum of stationary and continuous or fragmented and continuous are above 0.80, respectively. Time periods where none of these criterion are met are considered unclassified.

>>>>

Surprisingly, our results indicate that most events with coherent spatial content are not dominated by the continuous movement dynamic, but instead correspond to trajectories that are stationary or that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first note that most of the SWRs (14,989 of 21,433 or 70\% of classified SWRs) were classified as having only a single dynamic or dynamic mixture (\FIG{5}A, for specific example SWR see \FIG{3}B), whereas SWRs with multiple dynamics or dynamic mixtures (such as those in \FIG{3}A, C, D, E, and F) were less common. Interestingly, the most common category of SWR were those with only stationary-continuous mixtures (8,944 of 21,433 or 42\% of classified SWRs, \FIG{5}A, \FIGSUPP[5]{fig5sup1}). These events contain representations that move at slower speeds (\FIG{5}D)---similar to those speeds experienced by the animal in the environment (\FIG{5}G, median 17 cm/s for run periods and 4 cm/s for all times)---and are sustained, but slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 73 ms vs. continuous 94 ms, \FIG{5}B). Nonetheless, both the slow-moving events and continuous events frequently represent locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 52 cm vs. continuous 43 cm, \FIG{5}C). This indicates that the content of these SWRs, like those that are classified as continuous, do not represent the animal's actual position. We note that roughly a third of these events persisted for the entire duration of the SWR, but the other two thirds included some time where the model was uncertain about the dynamic. This shows the power of our state space decoder because it allows us identify periods of time when the model has high confidence with each SWR, and use these periods to characterize the SWR overall or to focus on specifically for further analysis.

@@

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with slower speeds of less than 1 m/s (<100 cm/s), which is consistent with the speeds that rats experience their environment. The next most common category were stationary events \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019} that activated a persistent representation of a single location, while the "classical" replay events---continuous, extended trajectories through space---made up only about 19\% of classified events. Events with slower dynamics tended to persist for approximately the same length of time as events  with faster continuous dynamics (\texttildelow50-100 ms), consistent with a brief activation of a representation of a location or a short trajectory through space. Events with slower dynamics most often represented the animal's current location, similar to continuous events identified in previous work \citep{DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}; however, slower events were also capable of representing locations far from the animal. Interestingly, Louie and Wilson \citeyearpar{LouieTemporallyStructuredReplay2001} found that during REM sleep, the longer time scale population activity (on the order of seconds) is highly correlated with firing during times when the animal is running, resembling the animal's experience. Our results are similar but on the timescale of SWRs, which last 10s to 100s of milliseconds, and during awake behavior.

@@

Fourth, our model avoided using large temporal bins and does not make assumptions about the linear progression of events. Instead, the model allows for movement in any direction allowed by geometry of the environment, and further, still accounts for the uncertainty of the data. Because the model gives a moment-by-moment estimate of the position and dynamic, this also minimizes the effect of misidentifying the SWR start and end times. Combined with an acausal estimation of latent position, this approach allows us to identify the specific timing of transitions in dynamics at a fine timescale. In addition to our SWR-based analyses, this approach could also be used to decode activity during movement, where we might expect that place-specific activity would express combinations of stationary and continuous dynamics reflecting different speeds of representational movement during each cycle of the theta rhythm. Similarly, because our approach uses the same encoder-decoder approach as the standard decoder, it can be used to investigate sequence dynamics before experience \cite{DragoiDistinctpreplaymultiple2013}.

@@

\subsection*{Classification of Dynamics}
We used a threshold of 0.80 to classify the probability of each state $Pr(I_{k} \mid O_{1:T})$ into 5 categories. Time periods during sharp wave ripples are labeled as stationary, continuous, or fragmented when the probability of each individual state is above 0.80, regardless of duration. Time periods are labeled as stationary-continuous-mixture or fragmented-continuous-mixture when the sum of stationary and continuous or fragmented and continuous are above 0.80, respectively. Time periods where none of these criterion are met are considered unclassified.
